### 2016-03-18 mcmc.tex
note to self
words on convergence and a new problem 9

% to-do
% -----
% - Deal with all mentions of ``DFM'' or ``HOGG''
% - Go through github issues, if any.
% - Systematically cite chapters of Brooks \etal in each section.
% - Skim MacKay and make sure we cite him in any relevant place.
% - Mention & cite STAN in methods section
% - Need to mention case where there is an MC integral INSIDE the likelihood function
% - Need to mention testing the likelihood function, by, say, 1-d scans back and forth.
% - Would love to have the Note symbols to be links to the notes and the notes to have links back to the main text...!
% - Send to friendlies for comments
%   - Gelman
%   - Goodman
%   - Lang
%   - Rix
%   - Bovy
%   - Brewer
% - Kill all issues in this list, above, and then:
%   - Post on arXiv
%   - Blog about it

@@

The autocorrelation function is defined by
\begin{eqnarray}
C(\Delta) &\equiv& \frac{E[(\pars_k - \mu)\,(\pars_{k+\Delta} - \mu)]}{E[(\pars - \mu)^2]}
\label{eq:autocorrelationfn}\\
\mu &=& E[\pars]
\quad ;
\end{eqnarray}
that is, it is the covariance of points separated by $\Delta$, taken
relative to the variance.
In the above equation, the assumption is that the samples $\pars_k$ are
given in the order that they were generated by the \MCMC\ algorithm.
By construction
\begin{eqnarray}
C(0) &=& 1
\\
\lim_{\Delta\rightarrow\infty} C(\Delta) &=& 0
\quad ,
\end{eqnarray}
and by definition, the autocorrelation ``time'' $\tau$ is
\begin{eqnarray}
\tau = 2\,\sum_{\Delta = 0}^{\infty} C(\Delta)
\label{eq:autocorrelationtime}\quad .
\end{eqnarray}
%% DFM: IS THIS DEFINITION CORRECT OR IS THERE A FACTOR OF 2 OR SOMETHING? OR A SUM FROM -infty to +infty?
Heuristically, the autocorrelation time is the number of samples you
need to take before you get a new independent sample.
More precisely, the autocorrelation time is the factor by which you
have to reduce the number of samples when you compute the variance on
a sampling-based expectation value estimate.

DFM: Here or in a note, can you give an empirical estimator of the
autocorrelation time?  The naive estimator is good to give, and we
could also give a good estimator too.

That is, if the expectation value $E[q]$ is computed by the sampling
approximation given in \equationname~(\ref{eq:expectationvalue}),
there is an uncertainty on that expectation given by
\begin{eqnarray}
  \mbox{$E[q]$ uncertainty} &=& \left[\frac{\tau}{K}\,\Var[q]\right]^{1/2}
  \\
  &\approx& \left[\frac{\tau}{K^2}\,\sum_{k=1}^K \left(q(\pars_k) - E[q]\right)^2\right]^{1/2}
  \quad ,
\end{eqnarray}
where we are being a bit loose with terminology because this is the
uncertainty not on $E[q]$ but on our sampling-based \emph{estimator} for
$E[q]$, and the $E[q]$ inside the sum is that estimator too.
It is slightly off-topic here, but related to this, if you have a long
chain of length $K$ and with autocorrelation time $\tau$ and you thin
the chain down to every $\tau$th sample (instead of using all $K$
samples), you will get just as good sampling-based estimates as if you
had used all $K$ samples.

@@

Of course, the autocorrelation-time definition
(\ref{eq:autocorrelationtime}) does not by itself provide an
\emph{estimator} for the time.
Usually, we just take the naive estimator---the ratio of empirical
expectations implied by \equationname~(\ref{eq:autocorrelationfn}),
and then perform a naive sum to get the autocorrelation time.
However, almost all empirical estimators of the autocorrelation time
will end up being under-estimates (biased low), because the finite
length of the chain makes any subset seem more representative than it
truly is; indeed this is a sub-problem of the problem that convergence
can never be known empirically with good confidence.

@@

\begin{problem}\label{prob:estimatetau}
Write a piece of code that computes the empirical autocorrelation
function and the empirical autocorrelation time from a chain using the na\"ive method DFM:WHERE?  Run
this on the chain you obtained from \problemname~\ref{prob:MH}.
Plot the autocorrelation function you get and over-plot the
autocorrelation time as an indicator on the plot.  What do you
conclude?
\end{problem}

\begin{problem}
Look at the source code of \project{emcee} \citep{emcee}, and see if you can
re-code or extract the autocorrelation-time estimator called DFM:WHAT?
This code should take a \MCMC\ chain of parameter vectors as input and
return a vector of autocorrelation times as output.
\end{problem}

>>>>

% to-do
% -----
% - Deal with all mentions of ``DFM'' or ``HOGG''
% - Go through github issues, if any.
% - Systematically cite chapters of Brooks \etal in each section.
% - Skim MacKay and make sure we cite him in any relevant place.
% - Mention & cite STAN in methods section
% - Need to mention case where there is an MC integral INSIDE the likelihood function
% - Need to mention testing the likelihood function, by, say, 1-d scans back and forth.
% - Make sure we say ``don't use MCMC for 1-d problems or if you have alternatives to directly integrate!''
% - Would love to have the Note symbols to be links to the notes and the notes to have links back to the main text...!
% - Send to friendlies for comments
%   - Gelman
%   - Goodman
%   - Lang
%   - Rix
%   - Bovy
%   - Brewer
% - Kill all issues in this list, above, and then:
%   - Post on arXiv
%   - Blog about it

@@

The autocorrelation function is defined by
\begin{eqnarray}
C(\Delta) &\equiv& \frac{E[(\pars_k - \mu)\,(\pars_{k+\Delta} - \mu)]}{E[(\pars - \mu)^2]}
\label{eq:autocorrelationfn}\\
\mu &=& E[\pars]
\quad ;
\end{eqnarray}
that is, it is the covariance of points separated by $\Delta$, taken
relative to the variance.
In the above equation, the assumption is that the samples $\pars_k$ are
given in the order that they were generated by the \MCMC\ algorithm.
By construction
\begin{eqnarray}
C(0) &=& 1
\\
\lim_{\Delta\rightarrow\infty} C(\Delta) &=& 0
\quad ,
\end{eqnarray}
and by definition, the autocorrelation ``time'' $\tau$ is
\begin{eqnarray}
\tau = 1 + 2\,\sum_{\Delta = 1}^{\infty} C(\Delta)
\label{eq:autocorrelationtime}\quad .
\end{eqnarray}
Heuristically, the autocorrelation time is the number of samples you
need to take before you get a new independent sample.
More precisely, the autocorrelation time is the factor by which you
have to reduce the number of samples when you compute the variance on
a sampling-based expectation value estimate.

That is, if the expectation value $E[q]$ is computed by the sampling
approximation given in \equationname~(\ref{eq:expectationvalue}),
there is an uncertainty on that expectation given by
\begin{eqnarray}
  \Var[E[q]] &=& \frac{\tau}{K}\,\Var[q]
  \\
  &\approx& \frac{\tau}{K^2}\,\sum_{k=1}^K \left(q(\pars_k) - E[q]\right)^2
  \quad ,
\end{eqnarray}
where we are being a bit loose with terminology because this is the
uncertainty not on $E[q]$ but on our sampling-based \emph{estimator} for
$E[q]$, and the $E[q]$ inside the sum is that estimator too.
It is slightly off-topic here, but related to this, if you have a long
chain of length $K$ and with autocorrelation time $\tau$ and you thin
the chain down to every $\tau$th sample (instead of using all $K$
samples), you will get just as good sampling-based estimates as if you
had used all $K$ samples.

@@

Of course, the autocorrelation-time definition
(\ref{eq:autocorrelationtime}) does not by itself provide an
\emph{estimator} for the time.
We might first try the na\"ive estimator
\begin{eqnarray}\label{eq:tauestnaive}
\tau &\stackrel{?}{=}& 1 + 2\,\sum_{\Delta = 1}^{N} C(\Delta)
\end{eqnarray}
where the sum is over the autocorrelation function estimated based the chain
using \equationname~(\ref{eq:autocorrelationfn}).
In practice, this is a high variance estimator of the integrated
autocorrelation time and this variance doesn't go to zero for an infinitely
long chain.\note{%
  A.~Sokal has a clear discussion of this topic and several proposals for
  robust autocorrelation time estimation in a set of lecture notes
  \url{http://www.stat.unc.edu/faculty/cji/Sokal.pdf}.}
Instead, we can get a much more robust estimate of the autocorrelation time by
restricting the sum in \equationname~(\ref{eq:tauestnaive}) to $\Delta \le M$
where
$\tau \le M \ll N$
\begin{eqnarray}\label{eq:tauest}
\tau &=& 1 + 2\,\sum_{\Delta = 1}^{N} C(\Delta)
\end{eqnarray}
In practice, $M$ can be chosen iteratively to satisfy these constraints but it
will only be as reliable as the chain itself.
An estimate of $\tau$ using \equationname~(\ref{eq:tauest}) should only be
trusted if the chain is many times ($\sim 10$) longer than $\tau$.

@@

\begin{problem}\label{prob:estimatetau}
Write a piece of code that computes the empirical autocorrelation
function.
You can speed this computation up considerably by using a fast Fourier
transform.
Run this on the chain you obtained from \problemname~\ref{prob:MH}.
Plot the autocorrelation function you find at short lags ($\Delta < 100$).
\end{problem}

\begin{problem}
Write a peice of code that estiamtes the integrated autocorrelation time for a
chain of samples using an estimate of the autocorrelation function and a given
``window'' size $M$ (see \equationname~\ref{eq:tauest}).
Plot the estimated $\tau$ as a function of $M$.
What can you conclude from this plot?
Implement an iterative procedure for automatically choosing $M$.\note{%
The recipe given on page 16 of A.~Sokal's notes
(\url{http://www.stat.unc.edu/faculty/cji/Sokal.pdf}) might be helpful.}
Overplot this estimate on the plot of $\tau(M)$.
\end{problem}
